---

  
---
# Project: Analyzing The Himalayan Mountain Range

## Data Description

My data source is from tidytuesdays and it is a set on Himalayan climbing expeditions (1905-2019).
The three datasets consist of Himalayan peak, expedition, and members information. We will only be focusing on the peak and expedition datasets.

My research questions that I hope to answer are:

  i) What peaks are the most popular among the Himalayas?
  ii)  What Himalayan peaks have the highest summit/survival rate? (Which mountains have been climbed the most with the least amount of deaths?)
  iii) What time of the year has the most successful summits? (No injuries/deaths)
  
## Data Cleaning

After loading in the data and before starting my analysis, I first added a column to the data set to denote if a climb was successful. After, I joined the peak and expedition dataset to create a singular dataset. I then created a few more variables, total expeditions, successful summits, death on treks, and then calculated the summit and death rates to use in my analysis.


```{r}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)
library(patchwork)
library(lubridate)

```


```{r}
#| label: load-in-data
#| echo: false
#| message: false
#| warning: false

#laoding in Tidy Tuesday Data
members <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-09-22/members.csv')
expeditions <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-09-22/expeditions.csv')
peaks <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-09-22/peaks.csv')


```

```{r}
#| label: cleaning-data
#| echo: false


#added columns denoting if climb was successful or if any deaths occurred
expeditions <- expeditions |>
  mutate(climb_success = case_when(
    termination_reason == "Success (main peak)" ~ 1,
    termination_reason == "Success (subpeak)" ~ 1,
    termination_reason == "Success (claimed)" ~ 1,
    .default = 0),
          death_on_exp = if_else(member_deaths + hired_staff_deaths > 0, 1, 0))
            
#joined peaks and expeditions dataset to add height of peaks, dropped unused columns         
peak_expeditions <- expeditions |>
  left_join(peaks |> select(peak_name, height_metres), by = c('peak_name')) |>
  select(-oxygen_used, trekking_agency)


peak_expeditions1 <- peak_expeditions |>
    group_by(peak_name) |>
  summarize(total_expeditions = n(),
            successful_treks = sum(climb_success),
            death_on_trek = sum(death_on_exp)) |>
  mutate(summit_rate = successful_treks/total_expeditions,
         death_rate = death_on_trek/total_expeditions) |>
  filter(total_expeditions >= 50) 


```

```{r}
#| label: write-data
#| eval: false
#| echo: false
write.csv(members, "data-raw/members.csv")
write.csv(expeditions, "data-raw/expeditions.csv")
write.csv(peaks, "data-raw/peaks.csv")
write.csv(peak_expeditions, "data-clean/peak_expeditions.csv")
write.csv(peak_expeditions1, "data-clean/peakexpeditions1.csv")

```


## Visualization 1

A rotated bar chart which displays the top 10 Himalayan mountains by total expeditions. The order of the mountains is in descending order by total expeditions. Mt. Everest claims the title for most popular, while Mt. Annapurna I is the least popular. 

```{r}
#| label: bar-plot-most-popular
#| echo: false
#| fig-cap: "The top 10 Himalayan mountains by total expeditions from 1905 to 2019"
#| fig-alt: "A bar chart showing the top 10 Himalayan mountains by total expedition from 1905 to 2019. It is ordered in descending order by total expeditions. Mt. Everest claims the title for most popular, while Mt. Annapurna I is the least popular."

total_treks <- peak_expeditions1 |>
  arrange(desc(total_expeditions)) |>
  head(10) |>
  #top 10 peaks by total # of expeditions
  ggplot(aes(y = fct_reorder(peak_name, total_expeditions), x = total_expeditions, fill = peak_name)) +
  geom_col() + # bar chart
  scale_fill_brewer(palette = "Spectral") +
  labs(
    title = "Most Popular Himalayan Peaks by Total Expeditions (1905-2019)",
    y = "Peak Name",
    x = "Number of Expeditions"
  )  +
  theme_bw() +
  theme(legend.position = "none") # remove legend
total_treks 
```

```{r}
#| label: bar-chart-summit-rate
#| echo: false
p1 <- peak_expeditions1 |>
  arrange(desc(total_expeditions)) |>
  head(10) |> #top 10 mountains by expedition
    ggplot(aes(y = fct_reorder(peak_name, summit_rate), x = death_rate, fill = peak_name)) +
  geom_col() +
  scale_fill_brewer(palette = "Spectral") +
  labs(
    y = "Name of Peak",
    x = "Death Rate"
  )  +
  theme_bw() +
  theme(legend.position = "none")
```


```{r}
#| label: bar-chart-death-rate
#| echo: false
p2 <- peak_expeditions1 |>
  arrange(desc(total_expeditions)) |>
  head(10) |> 
    ggplot(aes(y = fct_reorder(peak_name, summit_rate), x = summit_rate, fill = peak_name)) +
  geom_col() +
  scale_fill_brewer(palette = "Spectral") +
  labs(
    y = "Name of Peak",
    x = "Summit Rate"
  )  +
  theme_bw() +
  theme(legend.position = "none")
```

## Visualization 2

A side by side comparison of the summit rate (reaching the top of a mountain) and death rate. A ratio was constructed, as some mountains have a vastly higher number of total expeditions of others. Also because of this, we are only observing mountains among the top 10 most climbed. Both bar charts are ordered in descending summit rate, to show easy comparison of summits and deaths among the mountains. We can easily see that most mountains with a higher summit rate have a lower death rate. Observing Mt. Everest and Mt. Annapurna I again, we can see Everest among the top 3 in successful summit rate and death rate, while Annapurna I is dead last in summit and death rate.
```{r}
#| label: summit-death-comparison
#| fig-cap: "Side by side comparison of summit and death rate among the Himalayas"
#| fig-alt: "Two rotated bar charts that show the summit rate (successfully reach the top of a mountain) and the death rate among the top 10 most popular Himalayan peaks. Both bar charts are ordered in descending summit rate, to show easy comparison of summits and deaths. We can easily see that most mountains with a higher summit rate have a lower death rate. Observing Mt. Everest and Mt. Annapurna I again, we can see Everest among the top 3 in successful summit rate and death rate, while Annapurna I is dead last in summit and death rate."
#| echo: false
p2 + p1 + # side by side bar charts
  plot_annotation(
    title = "Summit Rate and Death Rate Among 10 Most Popular Himalayan Peaks (1905-2019)"
  )
```

## Visualization 3

This bar chart stacks the number of deaths on top of the number of successful summits in the Himalayan mountain range by month. We can easily see that the two most popular months are April and September, as the weather is less extreme. These most popular months claim the most successful summits and the most lives, showing that it is dangerous throughout the year to climnb in the Himalayas.
```{r}
#| label: deaths-and-summits
#| fig-cap: "Observing the deaths and summits among the Himalayan mountain range by month"
#| fig-alt: "A stacked bar chart that depicts the magnitude of number of climbs by month. The majority of expeditions are in April and September, when the weather is at its least extreme. We also see that the most popular months have the most successful summits and deaths, showing that there truly is no safe time of the year to climb in the Himalayas."
#| echo: false

top5_peaks <- peak_expeditions1 |> # obtaining the top 5 peaks to limit size of bars
  group_by(peak_name) |>           
  summarize(total = n()) |>        
  slice_max(total, n = 5) |>
  pull(peak_name)

peak_expeditions |> # using different data set that possesses date variable
  mutate(month = month(ymd(basecamp_date), label = TRUE, abbr = TRUE)) |>
  group_by(peak_name, month) |>
  summarize(total_expeditions = n(),
            success_treks = sum(climb_success), #constructing successful treks
            death_on_trek = sum(death_on_exp), .groups = "drop") |> # deaths on treks
    pivot_longer(cols = c(success_treks, death_on_trek), # pivoting longer to be able to stack
               names_to = "outcome",
               values_to = "count") |>
  ggplot(aes(x = month, y = count, fill = outcome)) +
  scale_fill_brewer(palette = "Spectral", name = "Death or Success",
                    labels = c("Death on Trek", "Summit Reached")) +
  geom_col() +
  labs(title = "Breakdown of Deaths and Summits by Month Amongst the Himalayas",
       subtitle = "The most popular months see the most successful summits and the most deaths",
       x = "Month",
       y = "Count") +
  theme_bw()

```



